\papertitle

### 2020-01-23 paper/paper.tex
Nai
Removed sections from NGI unneeded.

%\documentclass[a4paper, twocolumn]{article}
\documentclass[a4paper]{article}
\newcommand{\papertitle}{Whitepaper -- Optimising Storage Inside Workflows}

@@

\section{Workflows}

@@

@@

{ \color{blue}{

@@

For simplicity, we will call these datasets as \texttt{input-workflow.conf}.

@@

\item For simplification purposes, we will assume that Task\,1 has the datasets \texttt{input1.conf}, \texttt{output1.conf}, and \texttt{storage1.conf} as descriptors.

@@

\item Task\,2 has the datasets \texttt{input2.conf}, \texttt{output2.conf}, and \texttt{storage2.conf} as descriptors.

@@

\item Task 3 has the datasets \texttt{input3.conf}, \texttt{output3.conf}, and \texttt{storage3.conf} as descriptors.

@@

For simplicity, we will call these datasets as \texttt{output-workflow.conf}. In this specific example, it is composed by the datasets \texttt{output1.conf}, \texttt{output2.conf}, and \texttt{output3.conf}.

@@

For this work, the datasets belonging to \texttt{input-workflow.conf} are the datasets inserted as input for the running Cylc. The Cylc tool receives, as predefined, information about the storage and dependencies of the datasets. This information is used to construct and run the complete workflow, but it is not used to optimise intermediate steps. The configuration is already predefined, and it cannot change. For instance, if it is established that Task\,2 depends on Task\,1, then we have to wait for Task\,1 to be completed to start running Task\,2. But several options can be optimised if we know the whole workflow in advance. Thinking what about I/O, for instance, no decision is being made about an optimal way to store the data. Consider, as an example, that the datasets from \texttt{output1.conf} are divided into \texttt{output1-final.conf} and \texttt{output1-task2.conf}. Then, it is clear that \texttt{output1-task2.conf} will be required to run Task\,2, and, because of that, should be put in fast storage to quickly being retrieved, or even kept in local storage. The \texttt{output1-final.conf} dataset, however, can go directly to tape, if the lifetime of the final output has to be stored for a long time, but no access is required from the current workflow.

@@

\item[Low-level Interfaces]

In a nutshell, a stream of bytes is not a workable data model.
HPC data management and data-flow processing utilise low-level interfaces.
Storage relies still on traditional file system objects of directories and files -- provided by the POSIX interface or by object storage.
Underlying storage systems are receiving or delivering streams of bytes, and it cannot make any intelligent decisions about how to optimise data layout for performance, or resiliency.
It deals with data blocks and files and requires instructions on how to select and control available optimisation strategies.

In practice, applications build up layers so that the application itself is usually working with data mediated by a library which provides application-specific semantics. However, at the lowest level, such a library is limited by the underlying storage APIs (POSIX),
which has performance implications due to the synchronisation in distributed environments by severely restricting the ability of the underlying storage systems to optimise I/O operations.
Data-driven processing suffers similarly, from the high-level view of processes that are executed on a given computing node.

\item[Explicit Workflows]

Fair enough, batch systems understand the priority for jobs.
However, once a set of nodes is allocated explicitly to a computing job, the job runs exclusively, and it may utilise shared resources such as storage systems and network as any other job.

This problem is even more severe for data products: a storage system provides a single level of performance and resilience and performance for all use cases.
Theoretically, different fault-tolerance and performance policies can be realised with different storage systems and using management policies.
However, these policies would be coarse-grained and would not account for the priority of the workflow and, most importantly, the value of data.
The importance of data depends on aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow.
While value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability, in practice, this is not the case as the storage just sees blocks and files.

\item[Access Pattern in Transparency]

As access patterns are primarily implicitly defined in application code and embedded in workflows, the understanding and debugability of behaviour are limited.
Firstly, there are many alternative ways of programming data-intensive computation to achieve identical data products.
Secondly, many storages and computing interfaces are low-level.
Thus, data centres have often limited insight into the access patterns as well.
Moreover, some users are abusing the storage system in non-obvious ways, for example, to exchange notifications.
The understanding and improvement of access patterns are a limited and challenging insight into behaviour leads to suboptimal tuning.

Our vision for the problems described in the first sections is a new approach for data management and data-driven computation.
Our proposed model tackles the challenge of storage and data-flow computation in a holistic way accepting that data storage is only aiding the goal of generating insight.
Besides providing a typical abstract data model, our insight treats domain metadata\footnote{ in addition to that, we mean non-technical information about the data, for instance, the name of an experiment.}, workflows, and information lifecycle management (ILM) as first-class citizens allowing the storage to understand these concepts.
The user information will be exploited by smart components which aim to utilise capabilities of heterogeneous system landscapes.

To describe the features and benefits of the approach further, we sketch the user and system perspective in the following, as this highlights how we believe we should work in the 21st century.
The vision covers not only HPC but spans cloud, enterprise, and personalised IT.

\subsection{User Perspective}

\newcommand{\bnf}[1]{\textless #1\textgreater}

From the user's perspective, the interface and system match the core requirements and expectations, including reliability, availability, and usability.

\item[Data Model]

The data model is based on state-of-the-art solutions like HDF5+X, where X indicates an extension for ILM, workflows and processing.
However, it abstracts from data localisation and low-level concepts like files and directories and puts its focus on data organisation and data structures.
The integration into applications is shown in \Cref{fig:apilayers}.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\columnwidth]{layers-api}
  \caption{Layers with the proposed API}
  \label{fig:apilayers}
\end{figure}

Applications may utilise the proposed API directly, or utilise existing middleware\footnote{Which will not be able to use the full set of features if not extending its interface.}.
From the perspective of a developer of middleware or an application like web publishing, the tool provides a standard feature set that reduces the amount of code replication inside domain-specific solutions.

\item[Namespace]

Data is accessed using information from the domain metadata, i.e., similar to using methods of data indexing services.
The system will not be relying on a hierarchical namespace, but a dynamic mapping to the POSIX namespace can be provided to enable compatibility for POSIX applications.
For example, a user may decide to see a hierarchy like an experiment \bnf{Date}/\bnf{Name}/\bnf{Timestep}/\bnf{Task}/\bnf{Dataset}.
A core observation here is that, although the user may provide suggested namespaces, it is up to the system to place the data into the storage in an optimised way and return the information to the user transparently.

\item[Workflow Specification]

\item[Behavioral Contracts]

Besides the core characteristics, users expect the execution of their workflows under constraints (including cost and time constraints).
The system is always expected to show the expected characteristics and shield the user from system-specifics and undesired events (like faults or data loss).
The characteristics may improve over time, for instance, by expanding hardware capabilities, but the user will not have to change their workflow specification.
Still, for expert users, interfaces for introspecting system state and decision making can always be provided.

\item[Prescriptive Analysis]

Similarly to an SQL Explain statement, a toolset for users and administrators that allows exploring scenarios provides an unprecedented level of introspection. For instance, a prediction for runtime and execution costs of a workflow on a given system is now easier to be achieved.
Additionally, sharing the workflow description with vendors allows them to conduct predictions for runtime and costs for their solutions.
The ecosystem may also prescribe alternative systems to use in an open environment, e.g., this part of the workflow should run on data centre A and this part on cloud provider B.

\subsection{System Perspective}

An implementation of the proposed API exploits the user-supplied information and system characteristics intelligently to meet the user expectations and smartly use heterogeneous computing and storage landscapes.
It is expected that machine learning techniques will be applied to serve the individual aspects and evolve the systems over time.

\item[Awareness]

The system knows and understands user metadata and the overarching workflows (with the user intentions).
It is self-aware, knowing the nominal characteristics of the hardware and software components, continually comparing the system and workflow state with the expectations to perform mitigating actions to heal deviations.
At best, each hardware component would ship with a behavioural model that can be integrated hierarchically up to a global system model.
With behavioural models, we not only mean performance models but reliability and cost-models as well.

\item[Resource Management]

The system is in charge to manage and provision the available hardware resources autonomously.
Ultimately, it aims to utilise the individual characteristics of heterogeneous computing and storage landscapes by using intelligent scheduling.
Data placement and localisation is beyond mere tiering strategies; parts of data may be placed on particular storage technology or replicated in different representations across storage characteristics.
In addition to that, you are serving different access patterns and workflows at the same time.
Naturally, the system will replicate (expand) upon the availability of storage space while reducing redundancy under the shortage of space.

\item[Liquid Computing]

The system is in charge to manage and provision the available hardware resources autonomously and to assign computing and storage resources to tasks.
In addition to that, it may use, e.g., GPU, CPU, and in-network computing capabilities to serve pieces of a single workflow.
Moreover, as workflows stretch across data centres and over cloud and edge, an environment that schedules and dispatches work and storage efficiently is vital.
We call this concept \textbf{liquid computing} as the computing resources are dynamically provisioned and assigned while data flows through the system are processed.
Naturally, processing closer to the data sources is likely to improve system throughput.
The programming for the processing will be based on a high-level functional language providing primitives for data-flow computing and compiled into a virtual machine or machine code when the code is to be dispatched.
The capabilities of different hardware components may be limited, realising only a subset of the language which must be taken into account by the scheduler.

\item[Extensible Ecosystem]

The hardware and software landscape will be open, allowing to integrate solutions from different hardware and software vendors.
Connected systems may use different interfaces, including traditional POSIX interfaces, NoSQL, or distributed hash tables.
Imagine that a specific class of workflows performs inefficiently on a given infrastructure indicating that a crucial processing or storage capability is missing.
Adding new capabilities or resources to the system will allow the scheduler to use them in subsequent user-workflows speeding up the processing, what should be now as easy as adding an appliance to a business environment.

{ \color{blue}{

\section{Interaction with Cylc}

@@

>>>>

%\documentclass[a4paper, twocolumn]{article}
\documentclass[a4paper]{article}
\newcommand{\papertitle}{Potential of IO-Aware Workflows in Climate and Weather}

@@

\section{Workflows in Climate/Weather}

This is how workflows in climate/weather look like with Cylc.
Picture with XIOS software stack (julian' has that one.).

@@

\section{Vision for Workflows Considering IO}

What changes now in terms of specification?

\subsection{Potential Benefit}

@@

@@

For simplicity, we will call these datasets as \texttt{input.ds}.

@@

\item For simplification purposes, we will assume that Task\,1 has the datasets \texttt{input1.ds}, \texttt{output1.ds}, and \texttt{storage1.ds} as descriptors.

@@

\item Task\,2 has the datasets \texttt{input2.ds}, \texttt{output2.ds}, and \texttt{storage2.ds} as descriptors.

@@

\item Task 3 has the datasets \texttt{input3.ds}, \texttt{output3.ds}, and \texttt{storage3.ds} as descriptors.

@@

For simplicity, we will call these datasets as \texttt{output-workflow.ds}. In this specific example, it is composed by the datasets \texttt{output1.ds}, \texttt{output2.ds}, and \texttt{output3.ds}.

@@

For this work, the datasets belonging to \texttt{input-workflow.ds} are the datasets inserted as input for the running Cylc. The Cylc tool receives, as predefined, information about the storage and dependencies of the datasets. This information is used to construct and run the complete workflow, but it is not used to optimise intermediate steps.
The configuration is already predefined, and it cannot change. For instance, if it is established that Task\,2 depends on Task\,1, then we have to wait for Task\,1 to be completed to start running Task\,2.
But several options can be optimised if we know the whole workflow in advance. Thinking what about I/O, for instance, no decision is being made about an optimal way to store the data. Consider, as an example, that the datasets from \texttt{output1.ds} are divided into \texttt{output1-final.ds} and \texttt{output1-task2.ds}. Then, it is clear that \texttt{output1-task2.ds} will be required to run Task\,2, and, because of that, should be put in fast storage to quickly being retrieved, or even kept in local storage. The \texttt{output1-final.ds} dataset, however, can go directly to tape, if the lifetime of the final output has to be stored for a long time, but no access is required from the current workflow.

@@

\section{Extending Existing Workflows}

@@

\begin{description}
\item[Explicit Workflows]

This problem is even more severe for data products: a storage system provides a single level of performance and resilience and performance for stored files.
Theoretically, different fault-tolerance and performance policies can be realised with different storage systems and using management policies.
However, these policies would be coarse-grained and would not account for the priority of the workflow and, most importantly, the value of data.
The importance of data depends on aspects like costs to reproduce the data (can it be reproduced easily), the type of the experiment (test run, production run) and runtime constraints for the overall and potential crucial workflow.
While value and priority should influence fault-tolerance strategies and imply the quality of service for performance and availability, in practice, this is not the case as the storage just sees blocks and files.

Our vision for the problems described in the first sections is a new approach for data management and data-driven computation.

\begin{description}
\item[Workflow Specification]
